Supplementary Material: Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-15 manuscript/main.qmd
update fig 1 and 2
revisions

::: {#fig-1}
![](../figures/fig_1.png){width="15cm"}

@@

@@

# modern R2
R2_modern <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")

@@

::: {#fig-2}
![](../figures/fig_2.png){width="15cm"}

@@

::: {#fig-3}
![](../figures/fig_3.png){width="10cm"}

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient.
:::

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig. S2). This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S3).

::: {#fig-4}
![](../figures/fig_4.png){width="10cm"}

>>>>

::: {#fig-1}
![](../figures/fig_1_revised.png){width="15cm"}

@@

We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material.

@@

# modern R2
R2_modern_median <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")
R2_modern <- readRDS("results/modern/Rsquared_modern.rds")

@@

::: {#fig-2}
![](../figures/fig_2_revised_final.png){width="15cm"}

@@

::: {#fig-3}
![](../figures/fig_3_revised.png){width="10cm"}

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise ($R^2 = `r round(R2_modern,2)`, N = 42,896)$. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient. The darkness of the background shading indicates the point density.  
:::

@@

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude (black line) is explained by the modelled gradient ($`r 100*as.numeric(R2_modern_median)` \%$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median latitudinal gradient is higher near the equator and in high latitudes when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig. S2). This indicates that the ecological proxy data provide additional constraints on the shape of the early Eocene temperature gradient. In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S4).

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S4).

::: {#fig-4}
![](../figures/fig_4_revised.png){width="10cm"}

### 2023-09-15 manuscript/supplementary.qmd
revisions

@@

::: {#fig-S2}
![](../figures/SM/FigS_sd_included.png){width="10.5cm"}

@@

::: {#fig-S3}
![](../figures/SM/FigS3_north_south_eeco.png){width="12cm"}

Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue). Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data. The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
:::
<!-- REMOVED
```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credible intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided. If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
-->

>>>>

Estimated latitudinal temperature gradient using only the geochemical proxy data (yellow), showing the median (line) and 95% credible interval (shading). Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
:::

::: {#fig-S3}
![](../figures/SM/FigS_wider-ecol-limits.png){width="10.5cm"}

Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading). The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (@Kleypas1999), and a theoretical maximum derived by [@jones2022reef]. Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. The black line and shading in the background show the latitudinal temperature gradient with the narrower ecological limits, as in Figure 4.
:::

::: {#fig-S4}
![](../figures/SM/FigS3_north_south_eeco.png){width="12cm"}

Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue). Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$. Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data. The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres. The dotted lines show the empirical, modern gradient, averaged across bins of 1&deg, for both hemispheres combined (black), and for the Northern hemisphere (blue) and Southern hemisphere (red), separately.
:::

@@

::: {#fig-S5}
![](../figures/SM/FigS_sd_included.png){width="10.5cm"}

@@

<!-- REMOVED
```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4. Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data. The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credible intervals (shadings) of the inferred gradients using the model are shown in red. \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided. If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
setwd('..') # move up to parent folder of /manuscript
knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
```
-->
